Topic categorization in Portuguese news articles\\ \thanks{Fundação para a Ciência e Tecnologia}

### 2018-01-03 article.tex
Preprocessing

The simplest way of achieving this categorization is to have the
author of the article to manually introduce it (e.g. the journalist
typing it on the news article authoring framework); however, this
solution presents some challenges:
\begin{itemize}
    \item it increases the amount of work the author has to do
    \item the author might not be sure which categories are available
    \item the author might not be sure which category is the best
        (e.g. \emph{Economics} vs  \emph{Finance})
    \item it dows not scale -- e.g. if the goal is to categorize an
        existing (large) corpus
\end{itemize}

@@

\section{Methods}
\subsection{The dataset}
We gathered a dataset of news articles published in
\textit{Oservador}\footnote{\url{http://observador.pt}}, one of the
main Portuguese newspapers, and which stands out from the others for
being fairly young (it was created in May 2014) and for existing
exclusively online.

We gathered all the categories used by Observador, and ordered them
from the most common to the least common. We selected the ones which
had more than 1.000 articles in our dataset, and reduced our original
dataset to include only articles from these categories, which
were\footnote{Category names translated to English by the authors.}:
\begin{enumerate}
    \item International
    \item National
    \item Economics
    \item Politics
    \item Sports
    \item Science
    \item Culture
    \item Music
\end{enumerate}

@@

\begin{lstlisting}[caption={Example of JSON representation of an article},label={lst:article}, extendedchars=true]
{
  "Type" : "sapo.obj.creativework.article",
  "Source" : {
      "Name" : "Observador"
  },
  "Pretitle" : "Benfica",
  "Title" : "Ruben Amorim com rotura total do ligamento cruzado",
  "Author" : {
      "Name" : "Observador"
  },
  "Tags" : [
      "benfica",
      "desporto",
      "futebol",
      "ruben amorim"
  ],
  "PublishDate" : ISODate("2014-08-25T18:33:00Z"),
  "Lead" : "Depois de Fejsa, mais uma baixa. O internacional português [...]",
  "Body" : "<p>O pior cenário confirmou-se. O Benfica informou esta segunda-feira [...]",
  "URL" : "http://observador.pt/2014/08/25/ruben-amorim-com-rotura-total-ligamento-cruzado/",
  "CategoryPaths" : [
      "Desporto"
  ],
  "Domain" : "observador.pt",
  "Language" : "pt_PT",
  ...
}
\end{lstlisting}

@@

\subsection{Classification algorithms}
\subsubsection{Decision tree}
\subsubsection{k-nearest neighbors}
\subsubsection{Naive Bayes}
\subsubsection{Neural network}
\subsubsection{Support vector machine}

>>>>

The simplest way of achieving this categorization is to have the
author of the article to manually introduce it (e.g. the journalist
typing it on the news article authoring framework); however, this
solution presents some challenges:
\begin{itemize}
    \item it increases the amount of work the author has to do
    \item the author might not be sure which categories are available
    \item the author might not be sure which category is the best
        (e.g. \emph{Economics} vs  \emph{Finance})
    \item it does not scale -- e.g. if the goal is to categorize an
        existing (large) corpus
\end{itemize}

@@

In this article, we describe the preliminary results obtained in
developing a classifier to categorize news articles, and in developing
a similar classifier with restrictions added to the maximum size of
the lexicon used % TODO

\section{Methods}
%TODO
we first needed to choose and obtain a suitable dataset, and then
cleaning and preparing it to be used to train the classifiers.

\subsection{The dataset}
We gathered a dataset of news articles published in
\textit{Observador}\footnote{\url{http://observador.pt}}, one of the
main Portuguese newspapers, and which stands out from the others for
being fairly young (it was created in May 2014) and for existing
exclusively online. The initial dataset comprised 42.475 entries, from
which we used only a subset, for reasons later described.

We gathered all the categories used by Observador, and ordered them
from the most common to the least common. We selected the ones which
had more than 1.000 articles in our dataset, and reduced our original
dataset to include only articles from these categories.
Figure~\ref{fig:articles} presents an overview of the selected
categories and the number of articles available for each one.

\begin{figure}[htbp]
\centerline{\includegraphics[width=.97\columnwidth]{imgs/articles}}
\caption{Total number of articles retrieved for each category}
\label{fig:articles}
\end{figure}

@@

\begin{lstlisting}[caption={Example of JSON representation of an article},label={lst:article}, extendedchars=true]
{
  Type: "sapo.obj.creativework.article",
  Source: {
      Name : "Observador"
  },
  Pretitle: "Benfica",
  Title: "Ruben Amorim com rotura total do ligamento cruzado",
  Author: {
      Name: "Observador"
  },
  Tags: [
      "benfica",
      "desporto",
      "futebol",
      "ruben amorim"
  ],
  PublishDate: ISODate("2014-08-25T18:33:00Z"),
  Lead: "Depois de Fejsa, mais uma baixa. O internacional português [...]",
  Body: "<p>O pior cenário confirmou-se. O Benfica informou esta segunda-feira [...]",
  URL: "http://observador.pt/2014/08/25/ruben-amorim-com-rotura-total-ligamento-cruzado/",
  CategoryPaths: [
      "Desporto"
  ],
  Domain: "observador.pt",
  Language: "pt_PT",
  ...
}
\end{lstlisting}

@@

Originally, the dataset was obtained as large (more than 2.5 million
entries) MongoDB collection, containing articles from several
Portuguese and international newspapers. The process required to
transform this collection into data our classifiers could process
required: querying the database, exporting the news articles and
splitting them into a train and an evaluation datasets.

The database query selected articles from \texttt{Observador} where
the body had a length greater than 100 characters (to
discard some malformed articles which had an empty body or a body
composed of only a few words), and the categories included at least
one of the categories we previously selected.

For each article returned by the query, the pretitle, title, subtitle
and lead fields, if present, were simply copied to a plain text file,
separated by blank lines. The body field, however, was stored in the
database in HTML format. As such, the HTML tags had to be stripped,
and then it was also added to the plain text file.

The files were then stored in folders, separated by the category.
Furthermore, for each category, 700 articles were allocated to the
train set, and 200 to the evaluation set.

These preprocessing tasks were accomplished using Bash and Node.js
scripts.

\subsection{Classification algorithms}

\subsubsection{Decision tree}
\subsubsection{k-nearest neighbors}
\subsubsection{Naive Bayes}
\subsubsection{Neural network}
\subsubsection{Support vector machine}
